---
title: "ShinyRems Users Guide"
resource_files:
  - images/PRE_1.png
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
  
vignette: >
  %\VignetteIndexEntry{shinyrems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(shinyrems)
library(knitr)
library(grid)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = FALSE
)
img_path <- "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images"

```

```{r echo=FALSE}
library(png)
embed_png <- function(path, dpi = NULL) {
  meta <- attr(png::readPNG(path, native = TRUE, info = TRUE), "info")
  if (!is.null(dpi)) meta$dpi <- rep(dpi, 2)
  knitr::asis_output(paste0(
    "<img src='", path, "'",
    " width= '100%' />"
  ))
}
```


ShinyRems is an online application that allows a user to view, clean, plot and calculate simple statistics using water quality data from the B.C. government Environmental Monitoring System (EMS) database. 

## Using the App
- The shinyrems app is in a completed phase, however like any application bugs do occur. If the app greys out and/or freezes, reload the internet browser. If that doesn’t fix the issue, close the web browser, revisit RStudio and rerun the launch code to restart the app. 

- The app requires Chrome or Firefox internet browsers to run. If Internet Explorer is the users default internet browser, once the app opens in Internet Explorer, the user can copy and paste the web link into Chrome or Firefox.  

- Once the app opens in the users’ web browser, go through the tabs sequentially as each tab depends on the previous for information. The user can go back a tab to make changes if necessary, without restarting at tab 1.  

- In the app: variable = parameter, limit = guideline  

- A key feature of the app, and to keep track of the changes that occur to the dataset each tab, a data download option is available after each stage of data manipulation from raw, tidying, cleaning, outlier removal, plotting and summary statistics. 

- Blue question mark circles are present throughout the app to explain what each option is asking and how those choices will affect the dataset. 

## Getting Started
This app must be launched from the RStudio program. The app then opens in the users’ computer’s internet browser. Given the EMS database is so large, the app accesses EMS data cached on the users’ computer and why the app cannot be directly opened from a web link. 

### Download R and R Studio
The latest versions of the computer programs R and RStudio are essential to running the app. See [this website](https://github.com/bcgov/bcgov-data-science-resources/wiki/Installing-R-&-RStudio) for how to install the latest versions of R and RStudio on a BC government workstation.

### Launch ShinyRems App
Once R and RStudio are installed on the users’ computer, there are two options for getting the R code to launch the app:  

1. View the README at [this website](https://github.com/bcgov/shinyrems) to follow the instructions to launch the app. Copy and paste the code into a new R Script file within RStudio:
File -> New File -> R Script (Figure 1).  

2. Contact Heather Thompson (heather.c.thompson@gov.bc.ca) or Karly Harker (karly.harker@gov.bc.ca) for existing R scripts containing the code in Figure 1 to launch the app.  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_1.png" width="100%">
Figure 1: R code script in the RStudio program. This is the code to run to open the shinyrems app.

When initially launching the app, it is recommended to run each line of code by itself. Line 6 in Figure 1 will cache the latest versions of the EMS database onto the users’ computer. This will take awhile. Further help is at [this website](https://github.com/bcgov/rems) if needed.

There are six options for running the app, depending what year of EMS data is required:  

- demo’ - uses a demo dataset that requires no download of EMS data  

- ‘2yr’ - uses the most recent 2 years of EMS data  

- ‘4yr’ - uses the most recent 4 years of EMS data 

- ‘historic’ - uses the historic EMS data  (1965 - (Current Year - 2 Years))

- ‘all’ - uses combined “2yr” and “historic” EMS data  

- ‘upload’ - allows user to upload their own data following correct format. For details see “Using the upload data option” section below.  

See the BC Data Catalogue for more information about the [EMS datasets](https://catalogue.data.gov.bc.ca/dataset/bc-environmental-monitoring-system-results).


Pick the option required and type it into line 22 as shown in Figure 1 or below. Press the ‘Run’ button on the RStudio menu bar (Figure 1). The app will open in the users’ web browser. For example, the line of code below will launch the app with all EMS data. 
```
shinyrems::run_ems_app('all')
```

After running line 22 above, Figure 2 and Figure 3 show the questions asked in the RStudio console and require the users input. 
When the text “Listening on http…”, this means the app has launched in the users’ internet browser. 

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_2.png" width="100%">
Figure 2. A message to answer when the code to launch the app is run. The ‘2yr’ EMS dataset is updated every day.

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_3.png" width="100%">
Figure 3. A message to answer when the code to launch the app is run. This message confirms a data look up table can be stored on the users’ computer.


## ShinyRems App Features
### Tab 1: Data

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_4.png" width="100%">
Figure 4. The first screen seen when the app opens in the users web browser. The dataset chosen in the R code will be listed in the top left. Raw data can be chosen multiple ways, with options to view one parameter at a time and at a specific date range.

### Using 2yr, 4yr, historic or all options
When the app opens, Figure 2 is the initial app screen. The dataset chosen in the R code will be displayed in the top left of the screen. In the left-hand menu, the options for choosing the raw EMS dataset are:  

1. By permit number.   

2. By monitoring location or EMS ID. Multiple entries can be entered.  

3. The Site Map inset tab (Figure 3) can be used to find a monitoring location. Clicking a marker will add the site to the users list of selected sites. An option to zoom to Freshwater atlas watershed groups is available.  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_5.png" width="100%">
Figure 5. Raw data can also be found by selecting monitoring site locations on the Site Map inset tab.
 
One parameter can be chosen at a time for the site(s) selected.  In ANY of selected sites means that the parameters could be in one or more of the sites, in ALL the parameter list updates to what parameters are found in all the sites. A date range can be specified. Note this has bounds depending on the ‘2yr’, ‘4yr’ and/or ‘historic’ dataset chosen. The Get/Update Data button will need to be pushed if changes are made to the data input choices. The raw data from EMS appears in the inset Data tab.  Raw data is shown with a subset of 21 columns pertaining to water quality data:  
```
EMS_ID, MONITORING_LOCATION, LATITUDE, LONGITUDE, LOCATION_TYPE, COLLECTION_START, LOCATION PURPOSE, PERMIT, SAMPLE_CLASS, SAMPLE_STATE, SAMPLE_DESCRIPTOR, PARAMETER_CODE, PARAMETER, ANALYTICAL_METHOD_CODE, ANALYTICAL_METHOD, RESULT_LETTER, RESULT, UNIT, METHOD_DETECTION_LIMIT, MDL_UNIT, QA_INDEX_CODE, UPPER_DEPTH, LOWER_DEPTH.  
```

A download raw data button exists to keep track of the raw dataset. Make sure to name it something descriptive. 

### Using the upload data option

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_6.png" width="100%">
Figure 6. The first screen the user will see when the app opens in the users web browser when the ‘upload’ option was chosen.

If the user wants to upload a dataset not from EMS, choose the option ‘upload’:  
```
shinyrems::run_ems_app('upload')
```
When the app opens, Figure 4 is the screen the user will see. The user can download the template to view what format the data needs to be in (Figure 5). Data must be in a .csv file to upload into the app. Excel workbooks formats are not accepted. If the datafile is an excel file, export a worksheet to csv.
 
<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_7.png" width="100%">
Figure 7. Template of how the users’ data needs to be organized to upload into app.

## Tab 2. Tidy

The Tidy Tab manual options for tidying the dataset include (Figure 6):  

1. Select SAMPLE_STATE to include in dataset. The options present in the dataset will automatically show up. For more information see the Reference Tables Tab. Options include:  
    - Fresh Water                
    - Waste Water                 
    - HiVol                       
    - HiVol Selective Size Inlet  
    - Air Emission              
    - Dustfall                    
    - Stack                       
    - Soil                        
    - Animal Tissue               
    - Ground Water                
    - Refuse                     
    - Marine Water                
    - Plant Tissue  
  
2. Select SAMPLE_CLASS to include in dataset. The options present in the dataset will automatically show up. See the Reference Tables Tab for more information. Options include:  
    - Blank
    - Regular               
    - Replicate             
    - Replicate - Sequential
    - Replicate-First       
    - Replicate-Second          
    - Replicate - Paired  
    
3. MDL Action gives an option what to do with data points below the detection limit. Options are converting result value to zero, half, MDL or NA to leave as is. Note – next versions of the app will have a maximum likelihood estimation censored data model to correctly statistically deal with data below detection limits.  

4. Strict matching is an option to match exact parameter names to recognized parameter names. If this is not checked, the users’ dataset will include for example ‘Aluminum Dissolved’ and ‘Aluminum Total’, if the dataset from that EMS_ID includes both those parameters.  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_8.png" width="100%">
Figure 8. Tab 2 in ShinyRems. This tab is where the users raw data starts to be tidied and includes a few automatic and manual steps to customize the users dataset.

The Tidy Tab does a few things automatically:  

1. Standardizes EMS variables. It examines EMS parameter codes and converts all the same codes to the same variable name (Figure 7).  

2. Standardizes units. Units are changed to what the aquatic life water quality guideline is reported in.  

3. Columns are simplified to 15. The default columns retained are:  

    - EMS_ID
    - MONITORING_LOCATION (Renamed to "Station")
    - COLLECTION_START (Renamed to "DateTime")
    - PARAMETER (Renamed to "Variable")
    - PARAMETER_CODE (Renamed to "Code")
    - RESULT (Renamed to "Value")
    - UNIT (Renamed to "Units")
    - METHOD_DETECTION_LIMIT (Renamed to "DetectionLimit")
    - RESULT_LETTER (Renamed to "ResultLetter")
    - SAMPLE_STATE
    - SAMPLE_CLASS
    - SAMPLE_DESCRIPTOR

4. The timezone is set to PST.  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_9.png" width="100%">
Figure 9. The Messages inset tab of the Tidy Tab. This shows what is happening automatically in the Tidy Tab.

## Tab 3. Clean

The Clean Tab manual options for cleaning the dataset include (Figure 8):  

1. Remove blanks ensures all SAMPLE_STATE = Blank are removed from the dataset.  

2. Summarize by columns retains columns in the dataset. For example, if the user is assessing lake data, retaining UPPER_DEPTH and LOWER_DEPTH columns will likely be important to view data collected on the same day but at different depths. If the user simply wants to plot and summarize data by EMS_ID, just retain the EMS_ID column.  

3. Summarize by function gives an option of how the replicates can be summarized. Replicates are counted as two or more samples for the same variable on the same day. If the user has retained SAMPLE_CLASS = Replicate, these will be included with SAMPLE_CLASS = Regular in summarizing the same day samples. Options to summarize by ‘mean’, ‘median’, and ‘max’.  

4. Maximum CV is an option to specify the coefficient of variation between sample replicates. Replicates are counted as two or more samples for the same variable on the same day. Leave blank for the default value of infinity.  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_10.png" width="100%">
Figure 10. The main screen of the Clean Tab. This is where various options exist for cleaning the users now tidy dataset.
 
The Clean Tab does a few things automatically:
1. Replaces DateTime column with Date (YYYY-MM-DD).  

2. General description columns are removed from dataset:  
    - SAMPLE_STATE  
    - SAMPLE_DESCRIPTOR, SAMPLE_CLASS  
    - LOCATION_TYPE  

## Tab 4. Outliers

1. Standard deviations is an option to specify the number of standard deviations above which a data point is considered an outlier.  

2. Ignore values below detection limit is an option to specify whether to ignore data points below the detection limit when calculating the average deviation and calculating outliers.  

3. Large values only is an option to select whether only values which exceed the specified number of expected standard deviations should be identified as outliers. For example, any values less than the mean by the selected number of standard deviations will not be considered outliers.  

4. Remove outliers from plot. Choose whether to view the plot with or without outliers. Regardless, the outliers are removed from the dataset before moving onto the next tab.  

5. Plot customizations: Point size and plot height.  

6. Manual outlier removal: Click and drag mouse over a data point on the plot to manually select and remove outliers. This will remove it from the dataset.  

7. Plot legend: Outliers are identified by colour. Data are identified as Detected – True (above minimum detection limit) or Detected - False (below minimum detection limit; with a < in the Result Letter column).  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_11.png" width="100%">
Figure 11. The main tab of the Outlier screen. Options here include customizing what constitutes an outlier for the dataset and removing them if desired.

Final clean data with outliers removed can be downloaded as a .csv (Figure 10). 

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_12.png" width="100%">
Figure 12. The main screen of the Outliers Tab showing the Final Data inset tab. This shows the final clean dataset with outliers removed.

## Tab 5. Plots/Statistics

The Plots/Statistics tab allows for the initial visualization of the clean dataset with the addition of the aquatic life water quality guideline if desired, and a simple statistics summary.

Figure 11 shows the main Plot tab, with options for customization:  

1. Adjusting plot start and end date.  
2. Plot type: scatter or boxplot with options to show lines and/or points.  
3. Point size and line size.  
4. Facet by Variable to have the various EMS_IDs show up on the same plot, or Facet by EMS_ID to show one plot each for each EMS_ID.  
5. Colour by allows for options to colour by variable or EMS_ID.  
6. Plot height allows the plot to get larger.  
 
<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_13.png" width="100%">
Figure 13. The main plot screen. Here the user can view the initial visualization of the clean dataset with outliers removed. Various options exist for customizing the plot and downloading it to the users computer.

### Calculating aquatic life water quality guidelines
The app can calculate 170 freshwater aquatic life water quality guidelines (Figure 12). All the aquatic life guidelines and their details can be found under the ‘Limits’ drop-down menu on the Reference Tables tab. All BC approved water quality guidelines are also available on the [ENV guidelines website](https://www2.gov.bc.ca/gov/content/environment/air-land-water/water/water-quality/water-quality-guidelines/approved-water-quality-guidelines) and [DataBC](https://catalogue.data.gov.bc.ca/dataset/water-quality-guidelines-of-b-c-/resource/6f32a85b-a3d9-44c3-9a14-15175eba25b6).  

Notes on using the guideline feature:  

- It’s best to refer to the Limits table in the Reference Tables tab to determine what guidelines the app calculates and if it corresponds to the variable the dataset includes.  

- If the guideline is dependent on a parameter like chloride or hardness, the app first searches for any additional parameters required and then fetches all available data from EMS for that parameter within the timeframe/site that the user selected. All options selected by the user for cleaning/tidying the main dataset are reused for the dependent parameter data.  

- Select Term is an option to specify which aquatic life guideline for a specific variable to view on the plot. The options are ‘short’, ‘long’, and ‘long-daily’. Short = acute, long = long-term chronic and long-daily = long-term chronic but only 1 sample is needed, not 5 samples in 30 days.  

- Get modelled estimate models the guideline if there isn’t enough supporting information collected on the same day as the variable to calculate the guideline. If a guideline depends on another variable such as pH, Total Chloride, or Total Hardness, a parametric model is used to predict the pH, Total Chloride and Total Hardness for all dates with a value of any variable.  

- The function will fill in estimated values for dates for when Total Hardness, Total Chloride and pH haven't been collected but are required to determine the guideline. The model incorporates a seasonal element. If no concentration exists for the day of interest, then the dependent parameter concentration is assumed to be the average recorded value over 30 days. The model predicts the dependent variables and existing values are replaced. If there are less than 12 dependent variable values in all years of data, then an average value is taken. If there is only one year with 12 or more values, a simple seasonal smoother is used. If there is two years with 12 or more values, then a seasonal smoother with a trend is fitted. Otherwise a model with trend and a dynamic seasonal component is fitted.  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_14.png" width="100%">
Figure 14. The Guideline inset tab showing the options for calcuating the aquatic life guideline of the variable selected in the 1. Data tab.

There is also an option to manually enter a value (Figure 13). This could be a permit limit, water quality objective, another water quality guideline or another benchmark. Both the ‘set manually’ and ‘calculate from data’ option appear as a dashed line on the plot.  

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_15.png" width="100%">
Figure 15. The Guideline inset tab showing the option to set the guideline manually.

## Summary Statistics
The Summary Statistics inset tab on the Tab 5. Plot/Statistics gives a simple stats summary for each EMS_ID based on the clean dataset with outliers removed (Figure 14). Options include ‘n’, ‘min’, ‘max, ‘mean’, ‘median’, ‘standard deviation’, and ‘standard error’. This table can be downloaded as a .csv.

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_16.png" width="100%">
Figure 16. The inset Summary Statistics tab in the main Tab 5. Plot/Statistics. Each EMS_ID will represent a row.

## Tab 6: Reference Tables

Tab 6 shows the various EMS reference tables. These include information for each EMS column and its respective options. These tables will be helpful in Tab 2, 3 and 5 as the dataset is refined and guidelines are calculated. 

<img src= "https://raw.githubusercontent.com/bcgov/shinyrems/master/vignettes/images/PRE_17.png" width="100%">
Figure 17. The Reference Tables tab which show all options for the various EMS columns. These show the EMS reference tables available from Data BC.

## Resources

- [Water quality guidelines](https://www2.gov.bc.ca/gov/content/environment/air-land-water/water/water-quality/water-quality-guidelines/approved-water-quality-guidelines)  

- [Water quality objectives](https://www2.gov.bc.ca/gov/content/environment/air-land-water/water/water-quality/water-quality-objectives) (can be entered into the app in the ‘set manually’ option)  

- [Analyzing water quality data resources](https://www2.gov.bc.ca/assets/gov/environment/natural-resource-stewardship/nr-laws-policy/risc/guidlines_for_interpreting_water_quality_data.pdf)  

 
- [EMS training and manual](https://gww.nrs.gov.bc.ca/env/emre/welcome-ems-training-homepage) (intranet)  
 
- [Surface water monitoring sites map](https://governmentofbc.maps.arcgis.com/apps/webappviewer/index.html?id=0ecd608e27ec45cd923bdcfeefba00a7)

- [Water data and tools webpage](https://www2.gov.bc.ca/gov/content/environment/air-land-water/water/water-science-data/water-data-tools)

